### 2019-04-30 reports/article.tex
version submitted to biorxiv/pone

% title goes here:
\begin{flushleft}
{\Large
\textbf\newline{Spatiotemporal patterns of urbanization in three Swiss urban agglomerations: insights from landscape metrics, growth modes and fractal analysis}
}
\newline
% authors go here:
\\
Mart\'i Bosch\textsuperscript{1*},
J\'er\^ome Chenal\textsuperscript{1},
St\'ephane Joost\textsuperscript{2,1},
\\
\bigskip
\textbf{1} Urban and Regional Planning Community (CEAT) \'Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Lausanne, Switzerland
\\
\textbf{2} Laboratory of Geographic Information Systems (LASIG) \'Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Lausanne, Switzerland
\\
\bigskip
* Corresponding author: marti.bosch@epfl.ch

@@

\section*{Abstract}
% Urbanization is currently a global phenomenon that has become the most important form of landscape change and is increasingly affecting biodiversity and ecosystem functions. In order to evaluate the impacts of urbanization and inform urban planning, it is important to understand the spatiotemporal patterns of land use change associated to urbanization. This paper exploits three different frameworks, namely landscape metrics, urban growth modes and fractal analysis to characterize the spatiotemporal patterns of urbanization of the Swiss urban agglomerations of Zurich, Bern and Lausanne. The land use inventory provided by the Swiss Federal Statistical Office were used to assemble four temporal snapshots from 1980 to 2016 at the extent of the urban agglomerations. The time series of landscape metrics generally supports the diffusion and coalescence hypothesis, yet many remaining natural patches hinder the coalescence of urban patches.
% On the other hand, an analysis of the urban growth modes suggest that leapfrog development occurs at all periods, which contributes to an increasing fragmentation of natural patches and maintains the fractal configuration of the landscape.
% The discussion reviews potential explanations for the observed landscape changes, and concludes with some planning implications.
Urbanization is currently a global phenomenon that has become the most important form of landscape change and is increasingly affecting biodiversity and ecosystem functions.
In order to evaluate the impacts of urbanization and inform urban planning, it is important to understand the spatiotemporal patterns of land use change associated to urbanization.
This paper exploits three different frameworks, namely landscape metrics, urban growth modes and fractal analysis to characterize the spatiotemporal patterns of urbanization of the Swiss urban agglomerations of Zurich, Bern and Lausanne.
The land use inventory provided by the Swiss Federal Statistical Office were used to assemble four temporal snapshots from 1980 to 2016 at the extent of the urban agglomerations.
The time series of landscape metrics generally supports the diffusion and coalescence model of urban growth, with Zurich exhibiting most characteristics of coalescence while Bern and Lausanne seem to be at the transition between diffusion and coalescence.
Nevertheless, the analysis of the urban growth modes suggest that leapfrog development occurs at all periods, which contributes to an increasing fragmentation of natural patches and maintains the fractal configuration of the landscape.
The discussion reviews potential explanations for the observed landscape changes, and concludes with some planning implications.

@@

Recent decades have witnessed an increasing number of studies of the spatiotemporal patterns of land use change associated to urbanization \cite{dietzel2005spatio, seto2005quantifying, schneider2008compact, jenerette2010global, wu2011quantifying, li2013quantifying, liu2016general, nong2018quantifying}.
% hypotheses:
%% diffusion and coalescence, analogous to multiple nuclei development and envelopment
%% challenged such hypotheses  and concluded
% Initial comparative studies
Initial attempts to synthesis suggested that urbanization can be characterized as a two-step alternating process of diffusion and coalescence \cite{dietzel2005spatio, schneider2008compact}, nonetheless, subsequent studies challanged the empirical validity of such hypothesis.
% sample of 120 global cities from 1990 to 2000.
On the one hand, Jenerette and Potere \cite{jenerette2010global} examined the spatiotemporal patterns of land use change of a sample of 120 global cities from 1990 to 2000, and determined that overall, urbanization leads to fragmented landscapes with more complex and heterogeneous structures. On the other hand, Li et al. \cite{li2013quantifying} determined that the two-phase diffusion and coalescence model can be over-simplistic and that urbanization might be better characterized by means of three growth modes, namely infilling, edge expansion and leapfrogging, which operate simoultaneously while alternating their relative dominance.

@@

This paper intends to evaluate the spatiotemporal patterns of land use change observed in three of the largest Swiss urban agglomerations from three different perspectives, namely landscape metrics, urban growth modes and fractal analysis. 
The objective of this paper is twofold.
The first is to assess to test whether the spatiotemporal evolution of the three agglomerations conform to the diffusion and coalescence dichotomy, the second is to explore how the three adopted perspectives might complement each other.
The results will serve to discuss planning implications regarding the desirability of the recent densification policies adopted in Switzerland.
% The second is to explore how the such  might be related to characteristics of fractal objects.

@@

% The Federal Statute on regional planning of 1979 already included the responsibility to avoid sprawl by ensuring that land is used economically and that the extension of settlements must be limited ; also strengthened the role of the designated building zones and clearly reduced the number of new buildings constructed outside them.
% However, not effective because the municipalities can designate new building zones almost entirely autonomously
% Countryside initiative (August 2008) to improve existing regional planning law in two respects: (i) federal gov must promote high-quality internal urban development, and (ii) freeze total designated building zones for 20 years.
% Result: revision of the Federal Statute with (i) designation of building zones limited on anticipated need based on population growth forecast of the next 15 years, and (ii) levies to compensate for increase of property values. Accepted by Swiss voters in March 2013 and countryside withdrawn.
% Distribution of power:
%% federal: framework legislation and coordination
%% cantons: spatial planning (enact laws and comprehensive plans)
%% municipalites: land use plan and building ordinance ; compulsory since the Federal Statute of 1979
In line with the country's federalist government structure, the Swiss spatial planning system is distributed between the federal state, the 26 cantons and 2495 municipalities. The federal state specifies the framework legislation and coordinates the spatial planning activities of the cantons, while cantons check the compliance of municipal development plans with cantonal and federal laws. Municipal administrations are in charge of their local development plans, namely the land use plan and building ordinance, and might therefore be viewed as the most important spatial planning entities.
% Important aspects of spatial planning that are defined within federal laws include the delimitation of the number of new buildings constructed outside the building zone or the definition of quotas of arable land to be maintained for each canton.
While the Federal Statue on regional planning of 1979 limited the number of new buildings constructed outside the building zones, built-up areas have since increased continuously, mainly because the municipalities can designate new building zones almost entirely autonomously \cite{jaeger2014improving}. A major revision of the Federal Statue was accepted in 2013, which limits the municipal competences to designate new building zones and encourages infill development and densification by means of tax incentives.
% Current urbanization trends suggest significant increases of urban land use demands over the next decades, mostly at the expense of agricultural land located at the fringe of existing urban agglomerations \cite{price2015future}.
Forecasts based on the current urbanization trends predict significant increases of urban land use demands over the next decades, mostly at the expense of agricultural land located at the fringe of existing urban agglomerations \cite{price2015future}.

Given that a significant part of the cross-border urban agglomerations of Geneva and Basel (the second and third largest in Switzerland)) lie beyond the Swiss boundaries \cite{sfso2014espace}, in order to ensure coherence of the land use/land cover data (see the section below), this study comprises only three of the five largest Swiss urban agglomerations, namely Zurich, Bern and Lausanne \cite{sfso2018city}.
As shown in \autoref{figures/population_change.pdf}, the three cities have undergone important population growth over the last 30 years, especially during the most recent years and at the agglomeration extent.
With a total population over 1.3 million and land area of 1305 km$^2$ (1038 hab/km$^2$), Zurich is the largest Swiss urban agglomeration. As a leading global city and one of the world's largest financial centers, Zurich has the country's largest airport and railway station, and also hosts the largest Swiss universities and higher education institutions. 
% ranking among the most liveable cities in the world.
Bern is the capital of Switzerland and fourth most populous urban agglomeration in Switzerland, with a total population of 410000 inhabitants and occupying a land area of 783 km$^2$ (531 hab/km$^2$). As the fifth largest Swiss urban agglomeration and the second most important student and research center after Zurich, Lausanne has a total population of 409000 inhabitants over a land area of 773 km$^2$ (537 hab/km$^2$). Given its larger population growth rate, Lausanne is likely to soon surpass Bern and become the fourth largest urban agglomeration in Switzerland.
% Zurich: 1021859 (1990), 1080728 (2000), 1191058 (2010), 1354140 (2017)
% Bern: 351084 (1990), 349157 (1990), 357668 (2010), 415785 (2017)
% Lausanne: 300280 (1990), 311441 (2000), 339,389 (2010), 415596 (2017)
% Densities: Zurich: 1037 Bern: 531, Lausanne: 537
Overall, the three urban agglomerations are characterized by a pervasive public transportation system and a highly developed economy, with a 85\% of the employment devoted to the tertiary sector.

@@

The SFSO land statistics datasets have been used to produce a series of categorical maps for each urban agglomeration and time period.
The inventory is stored within a relational database where each entry corresponds to the coordinates of a given pixel and multiple columns that feature its respective land use/land cover classes for each survey period. In order to process the data in an automated and reproducible manner, an open source reusable toolbox to manage, transform and export categorical raster maps has been developed in Python \cite{bosch2019swisslandstats}.
The boundaries of each urban agglomeration have been adopted from the definitions provided also by the SFSO \cite{sfso2014espace}, which comprise multiple municipalities and have been established in consideration of population density, proximity between centers, economic activities and commuting behavior.
As stated above, Geneva and Basel are excluded from this study because a significant portion of their urban agglomeration lies beyond the extent covered by the SFSO land statistics inventory, namely the administrative boundaries of Switzerland.
% \cite{bosch2019swisslandstats}.
% The evolution of the urban agglomeration
The spatiotemporal evolution of the urban footprint for the three selected urban agglomerations (i.e., Bern, Lausanne and Zurich) over the study period (i.e., 1980-2016\textsuperscript{\ref{fn:years}}) is displayed in \autoref{figures/landscape_plots.pdf}.

@@

% Several studies have undertaken the analysis of the spatio-temporal patterns of urbanization from the perspective of landscape ecology.
% TODO: (this in intro) Recent years have seen an increasing number of studies that employ landscape metrics to characterize the spatiotemporal patterns of LULC change associated to urbanization. % TODO: nameref to SI Table here
% While a plentiful collection of landscape metrics can be find in the literature, many of them are highly correlated with one another. As a matter of fact, \cite{riitters1995factor} found that the characteristics discerned by 55 prevalent landscape metrics could be reduced to only 6 independent factors. Accordingly, most of the reviewed studies employ a reduced number of metrics to characterize the spatial landscape patterns.
While a plentiful collection of landscape metrics can be find in the literature, many of them are highly correlated with one another. As a matter of fact, Riitters et al. \cite{riitters1995factor} found that the characteristics discerned by 55 prevalent landscape metrics could be reduced to only 6 independent factors.
On the other hand, landscape metrics can be very sensitive to the resolution and the extent of the maps. However, several metrics empirically exhibit consistent responses to changing scales that conform to predictable scaling relations \cite{wu2002empirical, wu2004effects}.
Based on such remarks, and in order to enhance comparability with other studies, ten landscape metrics have been selected for the present study, whose details are listed in \autoref{selected-metrics}.
While complying with the FRAGSTATS v4 definitions \cite{mcgarigal2012fragstats}, the landscape metrics have been computed with the open source library PyLandStats \cite{bosch2019pylandstats}.
Like in most of the related studies, the categorical maps have been reclassified into urban and natural classes, and the metrics have computed at the urban class level, namely aggregating their values across all the urban patches of the landscape. The only exceptions are the contagion (CONTAG) and Shannon's diversity index (SHDI), which might only be computed at the landscape level, namely considering all the landscape classes (urban and natural in the this study).
Pixels that correspond to land unavailable for development, such as water bodies, have been excluded from the computation of the metrics.

@@

\paragraph*{Area-radius scaling}
% On the one hand, the radial dimension, which describes the relationship between the built-up area of an urban agglomeration and the distance from the main city center, has been shown to empirically conform to the relationship:
The relationship between the built-up area of an urban agglomeration and the distance from the main city center has been shown to empirically conform to the relationship:

@@

\begin{figure}[!ht]
  \begin{adjustwidth}{-.4\textwidth}{0cm}
    \centering
    \includegraphics[width=\linewidth]{figures/size_frequency_distribution.pdf}
    \vspace{.5em}
    \caption[Size-frequency distribution of urban patches]{\label{figures/size_frequency_distribution.pdf}Size-frequency distribution of urban patches for the three urban agglomerations at each temporal snapshot. Each dot represents an observed patch, with the color code denoting the temporal snapshot to which the patch corresponds. The colored dashed lines represent the best maximum likelihood fit to a power-law distribution for all the patch sizes observed at their corresponding temporal snapshot. The insets display an histogram with logarithmic bins of the urban patch size distribution. See \nameref{code-size-frequency-distribution}.}
  \end{adjustwidth}
\end{figure}

@@

% testing powerlaw hypothesis
%% goodness-of-fit test
% ``Practically, bootstrapping is more computationally intensive and loglikelihood ratio tests are faster. Philosophically, it is frequently insufficient and unnecessary to answer the question of whether a distribution ‘‘really’’ follows a power law'' (powerlaw, page 5)
% RESULTS: truncated power law is the best overall fit, followed by a lognormal distribution and then the powerlaw. Nevertheless, the scaling range for the truncated power law is big.
%% Notes: 1. truncated vs full: No point on overfitting. 2. powerlaw vs lognormal: small patches do not die.
On the other hand, the size frequency distribution of urban patches can be approximated by a power-law (see \nameref{code-size-frequency-distribution}), i.e., a straight line in the log-log scale, with its slope reflecting the scaling exponent $\alpha$ of (\ref{eq:size-frequency}). The slopes of the fitted curves are very stable through time for the three urban agglomerations, although they display a slight tendence to decrease over time (note the smaller slope of the fit curves of latter years in \autoref{figures/size_frequency_distribution.pdf} and \nameref{code-size-frequency-distribution})
% Despite a slight tendence of the fitted slope to decrease over time (note the smaller slope of the fit curves of latter years in \autoref{figures/size_frequency_distribution.pdf}), which in consonance with the trend of the MPS, reflects the fact that urban patches are becoming larger,
Such a trend reflects how urban patches are becoming larger, which is in consonance with the increase of MPS observed for all the urban agglomerations.
% Nevertheless, the increase of MPS is counterbalanced by the emergence of new urban patches, which as suggested by the irregular trend of PD might happen at any period. Overall, the smaller size of new urban patches seemingly contributes to the stability of the size-frequency distribution of urban patches \cite{white2015modeling}.
Nonetheless, the overall increase of MPS is counterbalanced by the emergence of new urban patches of smaller size, which seemingly contribute not only to keeping the size-frequency distribution of urban patches as a power-law, but also to scaling exponents that remain very stable through time. This is relatively surprising, especially considering that PD --- the metric that reflects the number of urban patches --- and the dominance of leapfrog growth show the most irregular trends.

@@

@@

Firstly, the time series of landscape metrics generally support the diffusion and coalescence hypothesis, with Zurich, the largest Swiss urban agglomeration, exhibiting most characteristics of coalescence, whereas Bern and Lausanne are seemingly undergoing a transition between diffusion and coalescence. Additionally, the trend of SHDI (not explicitly considered in the diffusion and coalescence formulation of Dietzel et al. \cite{dietzel2005spatio}) suggests that the landscape is becoming increasingly diverse in land use. %, which is consistent with the findings of Wu et al. \cite{wu2011quantifying}.

@@

@@

Altogether, the only metric that does not conform to the diffusion and coalescence dichotomy in the present study is CONTAG, which should increase as patches coalesce, yet monotonic decreases are observed for the three urban agglomerations. Such behavior suggests that urban and natural patches are becoming increasingly disaggregated and interspersed, which might result from a combination of three factors.

@@

% As noted by \cite{white1993cellular}, a long line of investigations of the fundamental properties of complex self-organizing systems suggest that successful structural evolution of complex systems embedded in variable envirnoments, such as cities, is only possible within fractal configurations, where the structure of the system is capable of adapting to changes of all scales.
% TODO: 1/f noise

@@

% In fact, \cite{white1993cellular} employed a series of cellular automata models of land use change to explore the
% Although the pervasiveness of fractals and power-laws within cities and regions is supported by a bold body of evidence which , including the present paper
% Although a bold body of studies, including the present article, provide strong evidence of the pervasiveness of fractals and power-laws among land use patterns of cities and regions,
% Although a bold body of studies have shown that high-level models from the complexity sciences are capable of replicating the fractal aspects of land use patterns, many of the foundations of such models are inconsistent with the inherent characteristics of land use change \cite{li2000fractal}.
% % This corresponds to the `equifinality problem', which \citep{richardson2002methodological} defined as situations where ``there are potentially many non-overlapping qualitatively different explanations/models for any particular phenomena resulting from nonlinear interactions'' (page 2).

>>>>

% title goes here:
\begin{flushleft}
{\Large
\textbf\newline{Spatiotemporal patterns of urbanization in three Swiss urban agglomerations: insights from landscape metrics, growth modes and fractal analysis}
}
\newline
% authors go here:
\\
Mart\'i Bosch\textsuperscript{*},
J\'er\^ome Chenal
\\
\bigskip
Urban and Regional Planning Community (CEAT), \'Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Lausanne, Switzerland
\\
\bigskip
* Corresponding author: marti.bosch@epfl.ch

@@

\section*{Abstract}
% Urbanization is currently a global phenomenon that has become the most important form of landscape change and is increasingly affecting biodiversity and ecosystem functions. In order to evaluate the impacts of urbanization and inform urban planning, it is important to understand the spatiotemporal patterns of land use change associated to urbanization. This paper exploits three different frameworks, namely landscape metrics, urban growth modes and fractal analysis to characterize the spatiotemporal patterns of urbanization of the Swiss urban agglomerations of Zurich, Bern and Lausanne. The land use inventory provided by the Swiss Federal Statistical Office were used to assemble four temporal snapshots from 1980 to 2016 at the extent of the urban agglomerations. The time series of landscape metrics generally supports the diffusion and coalescence hypothesis, yet many remaining natural patches hinder the coalescence of urban patches.
% On the other hand, an analysis of the urban growth modes suggest that leapfrog development occurs at all periods, which contributes to an increasing fragmentation of natural patches and maintains the fractal configuration of the landscape.
% The discussion reviews potential explanations for the observed landscape changes, and concludes with some planning implications.
Urbanization is currently a global phenomenon that has become the most important form of landscape change and is increasingly affecting biodiversity and ecosystem functions.
In order to evaluate the impacts of urbanization and inform urban planning, it is important to understand the spatiotemporal patterns of land use change associated to urbanization.
This paper exploits three different frameworks, namely landscape metrics, urban growth modes and fractal analysis to characterize the spatiotemporal patterns of urbanization of the Swiss urban agglomerations of Zurich, Bern and Lausanne.
The land use inventory provided by the Swiss Federal Statistical Office was used to assemble four temporal snapshots from 1980 to 2016 at the extent of the urban agglomerations.
The time series of landscape metrics generally supports the diffusion and coalescence model of urban growth, with Zurich exhibiting most characteristics of coalescence while Bern and Lausanne seem to be at the transition between diffusion and coalescence.
Nevertheless, the analysis of the urban growth modes suggest that leapfrog development occurs at all periods, which contributes to an increasing fragmentation of natural patches and maintains the fractal configuration of the landscape.
The discussion reviews potential explanations for the observed landscape changes, and concludes with some planning implications.

@@

Recent decades have witnessed an increasing number of studies of the spatiotemporal patterns of land use change associated to urbanization \cite{dietzel2005spatio, seto2005quantifying, schneider2008compact, jenerette2010global, wu2011quantifying, li2013quantifying, liu2016general, nong2018quantifying}.
% hypotheses:
%% diffusion and coalescence, analogous to multiple nuclei development and envelopment
%% challenged such hypotheses  and concluded
% Initial comparative studies
Initial attempts to synthesis suggested that urbanization can be characterized as a two-step alternating process of diffusion and coalescence \cite{dietzel2005spatio, schneider2008compact}, nonetheless, subsequent studies challanged the empirical validity of such hypothesis.
% sample of 120 global cities from 1990 to 2000.
On the one hand, Jenerette and Potere \cite{jenerette2010global} examined the spatiotemporal patterns of land use change of a sample of 120 cities distributed throughout the world from 1990 to 2000, and determined that overall, urbanization leads to fragmented landscapes with more complex and heterogeneous structures. On the other hand, Li et al. \cite{li2013quantifying} determined that the two-phase diffusion and coalescence model can be over-simplistic and that urbanization might be better characterized by means of three growth modes, namely infilling, edge expansion and leapfrogging, which operate simoultaneously while alternating their relative dominance.

@@

This paper intends to evaluate the spatiotemporal patterns of land use change observed in three of the largest Swiss urban agglomerations from three different perspectives, namely landscape metrics, urban growth modes and fractal analysis. 
The objective of this paper is twofold.
The first is to assess to test whether the spatiotemporal evolution of the three agglomerations conform to the diffusion and coalescence dichotomy, the second is to explore how the three adopted perspectives complement each other.
The results will serve to discuss planning implications regarding the desirability of the recent densification policies adopted in Switzerland.
% The second is to explore how the such  might be related to characteristics of fractal objects.

@@

% The Federal Statute on regional planning of 1979 already included the responsibility to avoid sprawl by ensuring that land is used economically and that the extension of settlements must be limited ; also strengthened the role of the designated building zones and clearly reduced the number of new buildings constructed outside them.
% However, not effective because the municipalities can designate new building zones almost entirely autonomously
% Countryside initiative (August 2008) to improve existing regional planning law in two respects: (i) federal gov must promote high-quality internal urban development, and (ii) freeze total designated building zones for 20 years.
% Result: revision of the Federal Statute with (i) designation of building zones limited on anticipated need based on population growth forecast of the next 15 years, and (ii) levies to compensate for increase of property values. Accepted by Swiss voters in March 2013 and countryside withdrawn.
% Distribution of power:
%% federal: framework legislation and coordination
%% cantons: spatial planning (enact laws and comprehensive plans)
%% municipalites: land use plan and building ordinance ; compulsory since the Federal Statute of 1979
In line with the country's federalist government structure, the Swiss spatial planning system is distributed between the federal state, the 26 cantons and 2495 municipalities. The federal state specifies the framework legislation and coordinates the spatial planning activities of the cantons, while cantons check the compliance of municipal development plans with cantonal and federal laws. With some exceptions, municipal administrations are in charge of their local development plans, namely the land use plan and building ordinance, and might therefore be viewed as the most important spatial planning entities.
% Important aspects of spatial planning that are defined within federal laws include the delimitation of the number of new buildings constructed outside the building zone or the definition of quotas of arable land to be maintained for each canton.
While the Federal Statue on regional planning of 1979 limited the number of new buildings constructed outside the building zones, built-up areas have since increased continuously, mainly because the municipalities can designate new building zones almost entirely autonomously \cite{jaeger2014improving}. A major revision of the Federal Statue was accepted in 2013, % which limits the municipal competences to designate new building zones and encourages infill development and densification by means of tax incentives.
which limits the amount of building zones that municipalities can designate and encourages infill development and densification by means of tax incentives.
% Current urbanization trends suggest significant increases of urban land use demands over the next decades, mostly at the expense of agricultural land located at the fringe of existing urban agglomerations \cite{price2015future}.
Forecasts based on the current urbanization trends predict significant increases of urban land use demands over the next decades, mostly at the expense of agricultural land located at the fringe of existing urban agglomerations \cite{price2015future}.

Given that a significant part of the cross-border urban agglomerations of Geneva and Basel (the second and third largest in Switzerland) lie beyond the Swiss boundaries \cite{sfso2014espace}, in order to ensure coherence of the land use/land cover data (see the section below), this study comprises only three of the five largest Swiss urban agglomerations, namely Zurich, Bern and Lausanne \cite{sfso2018city}.
As shown in \autoref{figures/population_change.pdf}, the three agglomerations have undergone important population growth over the last 30 years, especially during the most recent years and at the agglomeration extent.
With a total population over 1.3 million and land area of 1305 km$^2$ (1038 hab/km$^2$), Zurich is the largest Swiss urban agglomeration. As a leading global city and one of the world's largest financial centers, Zurich has the country's largest airport and railway station, and also hosts the largest Swiss universities and higher education institutions. 
% ranking among the most liveable cities in the world.
Bern is the capital of Switzerland and fourth most populous urban agglomeration in Switzerland, with a total population of 410000 inhabitants and occupying a land area of 783 km$^2$ (531 hab/km$^2$). As the fifth largest Swiss urban agglomeration and the second most important student and research center after Zurich, the Lausanne agglomeration has a total population of 409000 inhabitants over a land area of 773 km$^2$ (537 hab/km$^2$). Given its larger population growth rate, Lausanne is likely to soon surpass Bern and become the fourth largest urban agglomeration in Switzerland.
% Zurich: 1021859 (1990), 1080728 (2000), 1191058 (2010), 1354140 (2017)
% Bern: 351084 (1990), 349157 (1990), 357668 (2010), 415785 (2017)
% Lausanne: 300280 (1990), 311441 (2000), 339,389 (2010), 415596 (2017)
% Densities: Zurich: 1037 Bern: 531, Lausanne: 537
Overall, the three urban agglomerations are characterized by a pervasive public transportation system and a highly developed economy, with a 85\% of the employment devoted to the tertiary sector.

@@

The SFSO land statistics datasets have been used to produce a series of categorical maps for each urban agglomeration and time period.
% The SFSO inventory is stored within a relational database where each entry corresponds to the coordinates of a given pixel and multiple columns that feature its respective land use/land cover classes for each survey period.
% In order to process the data in an automated and reproducible manner, an open source reusable toolbox to manage, transform and export categorical raster maps has been developed in Python \cite{bosch2019swisslandstats}.
In order to process the SFSO datasets in an automated and reproducible manner, an open source reusable toolbox to manage, transform and export categorical raster maps has been developed in Python \cite{bosch2019swisslandstats}.
The boundaries of each urban agglomeration have been adopted from the definitions provided also by the SFSO \cite{sfso2014espace}, which comprise multiple municipalities and have been established in consideration of population density, proximity between centers, economic activities and commuting behavior.
As stated above, Geneva and Basel are excluded from this study because a significant portion of their urban agglomeration lies beyond the extent covered by the SFSO land statistics inventory, namely the administrative boundaries of Switzerland.
% \cite{bosch2019swisslandstats}.
% The evolution of the urban agglomeration
The spatiotemporal evolution of the urban footprint for the three selected urban agglomerations (i.e., Bern, Lausanne and Zurich) over the study period (i.e., 1980-2016\textsuperscript{\ref{fn:years}}) is displayed in \autoref{figures/landscape_plots.pdf}.

@@

% Several studies have undertaken the analysis of the spatio-temporal patterns of urbanization from the perspective of landscape ecology.
% TODO: (this in intro) Recent years have seen an increasing number of studies that employ landscape metrics to characterize the spatiotemporal patterns of LULC change associated to urbanization. % TODO: nameref to SI Table here
% While a plentiful collection of landscape metrics can be find in the literature, many of them are highly correlated with one another. As a matter of fact, \cite{riitters1995factor} found that the characteristics discerned by 55 prevalent landscape metrics could be reduced to only 6 independent factors. Accordingly, most of the reviewed studies employ a reduced number of metrics to characterize the spatial landscape patterns.
While a plentiful collection of landscape metrics can be find in the literature, many of them are highly correlated with one another. As a matter of fact, Riitters et al. \cite{riitters1995factor} found that the characteristics discerned by 55 prevalent landscape metrics could be reduced to only 6 independent factors.
On the other hand, landscape metrics can be very sensitive to the resolution and the extent of the maps. However, several metrics empirically exhibit consistent responses to changing scales that conform to predictable scaling relations \cite{wu2002empirical, wu2004effects}.
Based on such remarks, and in order to enhance comparability with other studies, ten landscape metrics have been selected for the present study, whose details are listed in \autoref{selected-metrics}.
While complying with the FRAGSTATS v4 definitions \cite{mcgarigal2012fragstats}, the landscape metrics have been computed with the open source library PyLandStats \cite{bosch2019pylandstats}.
Like in most of the related studies, the categorical maps have been reclassified into urban and natural classes, and the metrics have computed at the urban class level, namely aggregating their values across all the urban patches of the landscape. The only exceptions are the contagion (CONTAG) and Shannon's diversity index (SHDI), which can only be computed at the landscape level, namely considering all the landscape classes (urban and natural in the this study).
Pixels that correspond to land unavailable for development, such as water bodies, have been excluded from the computation of the metrics.

@@

\paragraph*{Area-radius scaling}
% On the one hand, the radial dimension, which describes the relationship between the built-up area of an urban agglomeration and the distance from the main city center, has been shown to empirically conform to the relationship:
The relationship between the built-up area of an urban agglomeration and the distance from the main city center has been shown (e.g., Frankhauser \cite{frankhauser1994fractalite}) to empirically conform to relationships of the form:

@@

\begin{figure}[!ht]
  \begin{adjustwidth}{-.4\textwidth}{0cm}
    \centering
    \includegraphics[width=\linewidth]{figures/size_frequency_distribution.pdf}
    \vspace{.5em}
    \caption[Size-frequency distribution of urban patches]{\label{figures/size_frequency_distribution.pdf}Size-frequency distribution of urban patches for the three urban agglomerations at each temporal snapshot. Each dot represents an observed patch, with the color code denoting the temporal snapshot to which the patch corresponds. The colored dashed lines represent the best maximum likelihood fit to a power-law distribution for all the patch sizes observed at their corresponding temporal snapshot. The insets display an histogram with logarithmic bins of the urban patch size distribution. The plots have been produced with the Python package powerlaw \cite{alstott2014powerlaw}. See \nameref{code-size-frequency-distribution}.}
  \end{adjustwidth}
\end{figure}

@@

% testing powerlaw hypothesis
%% goodness-of-fit test
% ``Practically, bootstrapping is more computationally intensive and loglikelihood ratio tests are faster. Philosophically, it is frequently insufficient and unnecessary to answer the question of whether a distribution ‘‘really’’ follows a power law'' (powerlaw, page 5)
% RESULTS: truncated power law is the best overall fit, followed by a lognormal distribution and then the powerlaw. Nevertheless, the scaling range for the truncated power law is big.
%% Notes: 1. truncated vs full: No point on overfitting. 2. powerlaw vs lognormal: small patches do not die.
On the other hand, the size frequency distribution of urban patches can be approximated by a power-law (see \nameref{code-size-frequency-distribution}), i.e., a straight line in the log-log scale, with its slope reflecting the scaling exponent $\alpha$ of (\ref{eq:size-frequency}). The slopes of the fitted curves are very stable through time for the three urban agglomerations, although they display a slight tendence to decrease over time (note the smaller slope of the fit curves of latter years in \autoref{figures/size_frequency_distribution.pdf} and \nameref{code-size-frequency-distribution}).
% Despite a slight tendence of the fitted slope to decrease over time (note the smaller slope of the fit curves of latter years in \autoref{figures/size_frequency_distribution.pdf}), which in consonance with the trend of the MPS, reflects the fact that urban patches are becoming larger,
Such a trend reflects how urban patches are becoming larger, which is in consonance with the increase of MPS observed for all the urban agglomerations.
% Nevertheless, the increase of MPS is counterbalanced by the emergence of new urban patches, which as suggested by the irregular trend of PD might happen at any period. Overall, the smaller size of new urban patches seemingly contributes to the stability of the size-frequency distribution of urban patches \cite{white2015modeling}.
Nonetheless, the overall increase of MPS is counterbalanced by the emergence of new urban patches of smaller size, which seemingly contribute not only to keeping the size-frequency distribution of urban patches as a power-law, but also to scaling exponents that remain very stable through time. This is relatively surprising, especially considering that PD --- the metric that reflects the number of urban patches --- and the dominance of leapfrog growth show the most irregular trends.

@@

\subsection*{Testing hypothesis of urbanization patterns}

@@

Firstly, the time series of landscape metrics generally support the diffusion and coalescence hypothesis, with Zurich, the largest Swiss urban agglomeration, exhibiting most characteristics of coalescence, whereas Bern and Lausanne are seemingly undergoing a transition between diffusion and coalescence. Additionally, the trend of SHDI, which is not explicitly considered in the diffusion and coalescence formulation of Dietzel et al. \cite{dietzel2005spatio}, suggests that the landscape is becoming increasingly diverse in land use. %, which is consistent with the findings of Wu et al. \cite{wu2011quantifying}.

@@

% % Although th
% In view of the fractal (hence self-similar) traits observed over the three urban agglomerations, and considering the urban agglomerations of Bern and Lausanne have almost identical population size, it is quite natural to ask to what extent the distinctive spatiotemporal patterns of urbanization observed in Zurich might be attributable to its larger population size, which is more than three times as large.
% % % Like in many biological systems, many properties of cities have been reported to be power law functions of its population size \cite{bettencourt2007growth}.
% % % This is a central question of the study of complex systems, and in fact, allometric studies of cities have reported that like many physiological characteristics of biological organisms, many properties of cities are power law functions of its population size \cite{bettencourt2007growth}.
% Such allometric perspectives are central to understand the potential universality of the processes driving urbanization \cite{bettencourt2007growth,batty2008size}, and therefore whether urban landscapes can be expected to globally become more homogeneous \cite{jenerette2010global}.
% % However, such matter cannot be properly addressed in the present study due to its limited scope.
% However, due to its limited scope, the present study is not well suited to address such matters.
% The only related observation is that the largest urban agglomeration, i.e., Zurich, is the most dense and shows the higher PLAND, shape complexity (as indicated by ED, AWMFD and LSI) and CONTAG, which suggest that despite the higher abundance of urban patches, they refuse to coalesce.

@@

Altogether, the only metric that does not conform to the diffusion and coalescence dichotomy is CONTAG, which should increase as patches coalesce, yet monotonic decreases are observed for the three urban agglomerations. Such behavior suggests that urban and natural patches are becoming increasingly disaggregated and interspersed, which might result from a combination of three factors.

@@

\subsection*{Examining the continuous fragmentation}

% As reviewed by White et al. \cite{white2015modeling}, urban patches are expected to grow as cities and regions grow, which suggests that eventually urban patches coalesce and thus the small ones disappear. Yet this contrasts with the continuous fragmentation and the persistence of the power-law scaling of the urban patches observed in the present study.
% As urbanization unfolds, urban patches are expected to grow, which suggests that eventually urban patches coalesce and thus the small ones disappear. Yet this contrasts with the continuous fragmentation and the persistence of the power-law scaling of the urban patches observed in the present study.
As urbanization unfolds, urban patches are expected to grow, which suggests that eventually urban patches coalesce and thus the small ones disappear. Yet this contrasts with the continuous fragmentation of natural patches and the persistence of the power-law scaling of the urban patches observed in the present study.
% % As noted by \cite{white1993cellular}, a long line of investigations of the fundamental properties of complex self-organizing systems suggest that successful structural evolution of complex systems embedded in variable envirnoments, such as cities, is only possible within fractal configurations, where the structure of the system is capable of adapting to changes of all scales.
% % TODO: 1/f noise
% The seminal work of White and Engelen \cite{white1993cellular} experimented with various cellular automata models of land use change and showed that for an appropriate level of stochastic perturbation, the simulated cities quickly self-organize into a fractal structure where the size distribution of urban patches exhibits a power-law. Furthermore, their experiments showed that such structure emerged robustly despite very different initial conditions, and was maintained throughout the simulation.

@@

% % In fact, \cite{white1993cellular} employed a series of cellular automata models of land use change to explore the
% % Although the pervasiveness of fractals and power-laws within cities and regions is supported by a bold body of evidence which , including the present paper
% % Although a bold body of studies, including the present article, provide strong evidence of the pervasiveness of fractals and power-laws among land use patterns of cities and regions,
% % Although a bold body of studies have shown that high-level models from the complexity sciences are capable of replicating the fractal aspects of land use patterns, many of the foundations of such models are inconsistent with the inherent characteristics of land use change \cite{li2000fractal}.
% % % This corresponds to the `equifinality problem', which \citep{richardson2002methodological} defined as situations where ``there are potentially many non-overlapping qualitatively different explanations/models for any particular phenomena resulting from nonlinear interactions'' (page 2).
